%!TEX root = ../main.tex

\chapter{Programmable realtime unit}
\label{ch:pru}

This chapter will discuss the capabilities, and usage of the Programmable Realtime Unit (\textls{PRU}).

Even with realtime operating systems, no process can take up hundred percent time of a \cpu. There are essential OS management tasks, which handles the hardware. These management tasks need to be executed periodically.

\section{Motivational example}
\label{subsec:motivational_example}

This example illustrates the usability of an external control unit, if the system has rigorous timing criteria.

\begin{figure}[h]
	\centering
	\begin{tikzpicture}[
			label/.style={
				font=\footnotesize
			}
		]
		\node[squarednode] (A) [] {\rtos};
		\node[squarednode] (B) [right = 4cm of A] {Controller};

		\draw[thick,->] (A.east) -- ++(4,0) node[above,pos=0.12] (AB) {};
		\draw[very thick,->] (B.east) -- ++(4,0) node[midway,above,text width = 3cm] {To controlled unit};
		\timing at (AB) {6{1C 0.5C}C};
	\end{tikzpicture}
	\caption{A bit-banging solution for PWM}
\label{fig:example_pwm}
\end{figure}

\cref{fig:example_pwm} depicts the example. Our example consist of a computer (\rtos), and a controller. The \rtos{} component is sending a pulse width modulated signal (\pwm) to the controller. The controlling computer does not have an integrated \pwm{} controller, and the signal modulation must be done by software. This method of emulation of a hardware peripheral is called \emph{bit-banging}.

In use cases where the controller needs a very precise signal, because a small change in the width of the modulated signal cause, a big difference in the output of the controller. This specification would imply a controller with the smallest reachable jitter.

This example relies on the measurements of \citep{rt-sched-thesis}, which uses the same development platform this chapter will discuss. It is important to point out that the measurements were done on a soft real-time \os. Other commercial products which specializes in hard real-time (VxWorks, µC/OS-II) might reach better performance, but this chapter will reveal alternate solutions.

\subsection{Evaluation}

At PWM period of 50ms, the system was capable generating accurate control signal from software. At 10ms, the best achievable jitter was $1.32\%$, and at 1ms, $4.1\%$.
Because the computer is not totally dedicated to generate our control signal, jitter can happen. If the specified system requires low jitter, this solution may not be applicable to the design.

\section{Delegation of real-time functionality}

One possible solution for \cref{subsec:motivational_example} is to delegate this task to an external coprocessor (\cref{fig:example_external_pwm}).

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\node[squarednode] (A) [text width=1.5cm] {\rtos};
		\node[squarednode] (B) [below = 1.5cm of A] {Microcontroller};
		\node[squarednode] (C) [right = 2cm of B] {Controller};

		\draw[thick,->] (A) -- (B) node[midway, right] {Commands};
		\draw[thick,->] (B) -- (C) node[above,pos=0.12] (AB) {};
		\draw[very thick,->] (C.east) -- ++(3.7,0) node[midway,above] {To controlled unit};
		\timing[timing/slope=.25] at (AB) {3{0.8C0.4C}C};
	\end{tikzpicture}
	\caption{An external microcontroller as PWM signal source}
\label{fig:example_external_pwm}
\end{figure}

With this approach the design have an external component. This component, usually a microcontroller, is a completely separate unit. The communication between these units can be done with protocols designed to tolerate errors in timing e.g.\ the Inter-Integrated Circuit (\isc) protocol.

The only job of the microcontroller is to generate the corresponding controlling signal. This is a far more controlled environment, because microcontrollers don't usually run operating systems\,--and while interrupt handling might cause problems\,--the order of robustness is definitely higher than in case of an RTOS based solutions.

On the downside, the design have an additional external component which is an overhead in the design and programming process. The optimal solution would be to merge the two processors into one.

\section{Programmable realtime unit}

Instead of having an external controller (\cref{fig:ext_controller_hw}), this external coprocessor could be integrated into the integrated circuit of the central processing unit (\cref{fig:int_controller_hw}).

This can be beneficial because we need less external components, and because of the integrated property of this solution, the communication with the coprocessor is reliable.

Additionally the coprocessor can use some of the resources of the host \cpu{} which would be hardly accessed in an external configuration.

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\node[squarednode] (A) [] {Main computing unit};
		\node[squarednode] (B) [right = 4cm of A] {External controller};

		\draw[thick,<->] (A.east) -- ++(4,0) node[midway,above] (AB) {Commands \& Data};
	\end{tikzpicture}
	\caption{External coprocessor}
\label{fig:ext_controller_hw}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{tikzpicture}[
			innernode/.style={
				minimum height=1cm,
				text width=2.8cm,
				align=center
			}
		]

		\node[squarednode, innernode] (ECU) [] {Embedded controller};
		\node[squarednode, innernode] (MCU) [left = of ECU.west] {Central processing unit};
		\node[squarednode] (A) [ultra thick, fit=(ECU) (MCU), inner ysep=5mm, inner xsep=5mm] {};

		\node [above = 2mm of A.south] {\textbf{\large IC}};

		\draw [thick,<->] (ECU.west) -- (MCU.east);
		\draw [thick,<->] (MCU.north) -- ++(0, 1);
		\draw [thick,<->] (ECU.north) -- ++(0, 1) ;
	\end{tikzpicture}
	\caption{Integrated coprocessor}
\label{fig:int_controller_hw}
\end{figure}

\subsection{AM335x processor family}

The AM335x is one of the Texas Intruments' ARM Cortex-A8 based processor family~\citep{AM335x}. This thesis will focus on the AM3358 processor, because the BeagleBone Black \citep{BBB} has this \cpu. The BeagleBone Black is the reference system this work will use for taking measurements.

The AM335x processor family have four members (AM3359, AM3358, AM3357, AM3356) which have special unit called \emph{Programmable Realtime Unit and Industrial Communication SubSystem} (\pruss).

\subsection{AM335x \pruss}

A \pruss{} consists two \emph{Programmable Realtime Unit} (\pru). These units are Hardware architecture based microcontrollers which can be programmed by an application running on the central processing unit.

The \pru{} is not an accelerator next to the media processor, it's sole purpose is to offload hard real-time software components into these units. In this way, the designer may not need a real-time operating system, which can be a big tradeoff in performance, capabilities, design and license costs.

\needspace{10\baselineskip}
The basic properties of the AM335x \pruss{} (be warned, there are previous generations of \pruss, generations do not always indicated):
\begin{itemize}
	\item 2 \pru{} processor per \pruss{}:
	\begin{itemize}
		\item \SI{32}{\bit} Harward architecture, without pipelining
		\item \SI{200}{\mega\hertz} clock speed, all instruction executed in \SI{5}{\nano\second}
		\item \SI{8}{\kilo\byte} instruction memory (\iram)
		\item \SI{8}{\kilo\byte} data memory (\dram)
		\item \SI{12}{\kilo\byte} shared memory between the two processor
		\item Up to 17 input, and 16 output
	\end{itemize}
	\item An interrupt controller shared between the \pru{} cores
	\item Access to the system peripherals, e.g.: DRAM, GPIO
	\begin{itemize}
		\item System DRAM access is possible, but nondeterministic
		\item GPIO handling is direct, meaning I/O can be toggled with \SI{5}{\nano\second} resolution.\footnote{The sustainable maximal frequency achievable is \SI{50}{\mega\hertz}, because the toggle loop takes 4 instructions to execute.}
	\end{itemize}
\end{itemize}

\begin{figure}[h]
	\centering
	\begin{tikzpicture}[
			squarednode/.style={
				rectangle,
				draw=black,
				very thick,
				align=center,
				inner sep=0cm
			},
			main/.style={
				minimum width=13cm,
				minimum height=8cm
			},
			interconnect/.style={
				minimum width=12cm,
				text width=12cm,
				minimum height=1cm
			},
			sram/.style={
				minimum width=3.75cm,
				text width=3.75cm,
				minimum height=1cm
			},
			ram/.style={
				minimum width=1.75cm,
				text width=1.5cm,
				minimum height=1cm
			},
			pru/.style={
				minimum width=3.75cm,
				text width=3.5cm,
				minimum height=2cm
			}
		]

		\node[squarednode,main] (Box) at (0,1) {};
		\node[squarednode,below right,inner sep=3mm, fill = black!5] at (Box.north west) {\large\textbf{\pruss}};

		\node[squarednode,interconnect] (SharedRam) at (0,0) [] {Interconnect};

		\node[squarednode,sram] (IntC)  at (-4,-2) {Interrupt Controller};
		\node[squarednode,sram] (Peri)  at (4, -2) {Peripherals};

		\draw[thick, <->] (-4,-1.5) -- (-4,-0.5);
		\draw[thick, <->] (4, -1.5) -- (4, -0.5);

		\node[squarednode,sram, fill=black!10] (IntC)  at (-4,-4.5) {PRU0 I/O};
		\node[squarednode,sram, fill=black!10] (Peri)  at ( 4,-4.5) {PRU1 I/O};

		\draw[very thick, <->] (-4,-3) -- (-4,-4);
		\draw[very thick, <->] ( 4,-3) -- ( 4,-4);

		\node[squarednode,sram] (IntCon)  at (-4,2) {Shared RAM};
		\node[squarednode,ram] (Pru0IRAM) at (-1,2) {Instr. RAM};
		\node[squarednode,ram] (Pru0DRAM) at (1, 2) {Data RAM};
		\node[squarednode,ram] (Pru1IRAM) at (3, 2) {Instr. RAM};
		\node[squarednode,ram] (Pru1DRAM) at (5, 2) {Data RAM};

		\draw[thick, <->] (-4,1.5) -- (-4,0.5);
		\draw[thick, <->] (-1,1.5) -- (-1,0.5);
		\draw[thick, <->] (1, 1.5) -- (1, 0.5);
		\draw[thick, <->] (3, 1.5) -- (3, 0.5);
		\draw[thick, <->] (5, 1.5) -- (5, 0.5);

		\node[squarednode,pru] (PRU0) at (0, 3.65) {PRU0};
		\node[squarednode,pru] (PRU1) at (4, 3.65) {PRU1};

	\end{tikzpicture}
	\caption{Block diagram of the \pruss{}}
\label{fig:pruss_block}
\end{figure}

\subsection{Determinism of the \pru}

Texas Instruments' main goal with this coprocessor was to maximize the determinism of the processor, and the code running on it. This approach utilizes some unique design decisions following in the next points.

\subsubsection{Execution}

On the \pru, every instruction takes \SI{5}{\nano\second} to execute. This is an interesting design pattern, because on most architecture the execution cycle varies with instructions e.g.\ the \emph{JUMP} instruction always need more execution cycle than a simple addition. The \pru{} architecture is uniformial. This uniformity helps the static analysis of code, because there are no corner cases in the cycle counts.

There is no pipelining, and out-of-order execution, so the processor execute the machine code in the exact same order it was compiled.

\subsubsection{Interrupt handling}

There is no interrupt vector however the \pruss{} have an interrupt controller. The handling of interrupts is not hardware based like in other microcontrollers like the Microchip's PIC or the Atmel's AVR microcontrollers.

The default approach in interrupt handling is to interrupt the normal code execution, save the state of the normal execution (current program counter, and register values), and jump to an interrupt service routine (\textls{ISR}). After the ISR returns, the normal execution is restored, and continued.

On the \pru{}, there is no hardware based approach, the executed code must poll a specific register containing the flag bits of interrupt signals. By this way, the determinism of interrupt handling (despite the name \emph{interrupt} implies the interruption of normal execution flow) is fully handled by the executed code thus giving full control when these signals are processed.

\section{Firmware loader}

Because the \pru{} have it's own instruction and data memory, a driver is needed to load the compiled code inside it. This firmware loader is different in linux kernel 3.x and 4.x. The thesis will use the 4.x kernel version loader driver.

\subsection{Linux kernel v3.x}

In linux kernel v3.x, the firmware, and resource initialization is entirely the responsibility of the programmer. A user-space application using the \emph{prussdrv} \citep{PRU_PRUSSDRV} framework can load the compiled program into the instruction memory of the \pru, and handle various resources e.g.\ memory mapping or interrupt signals.

\subsection{Linux kernel v4.x}
\label{subsec:fw_loader_kernel4}

From linux kernel 3.x to 4.x the entire framework structure changed. While 3.x version was completely custom, \ti{} developers moved forward to more standard linux framework like \emph{remoteproc} \citep{RPROC}. While remoteproc is harder to understand, it is a well documented framework.

The biggest change from kernel 3.x is the move to kernel module level. Instead of writing a user-space application, the kernel loading is done by a kernel module. The kernel module is called \emph{pruss-remoteproc}, and is responsible of loading the kernel bundled into an Executable and Linkable Format (\elf{}) file inside the memory of the \pru. Because the loading is placed into kernelspace, and the module which loads it is uniform and not written by the programmer, some configuration method is needed. This method is described in the \cref{subs:linking}

\section{Toolchain}

Because it's a custom environment, the compiler toolchain utilizes some specific approaches. This section will describe the biggest differences from an everyday used \verb/g++/ or \verb/clang/ compilation workflow.

\subsection{Compiler support}

Compilation to the \pru{} cores can be either be done by:
\begin{itemize}
	\item Assembly \citep{PRU_ASM}. Because the PRU is a RISC architecture, the number of instructions are rather limited supporting easy programming by assembly language \citep{PRU_ASM_INSTR}. The compiler for assembly files is \verb+pasm+.
	\item \cpl{} or \cpp{} language \citep{PRU_C_CPP}. The compiler for assembly files is \verb+clpru+. The compiler supports \cpl{99} and \cpp{03} versions of the languages. This compiler is specifically designed for compiling \cpl{}/\cpp{} to the \pru.
\end{itemize}
This thesis will use the \verb+clpru+ compiler, because the generated code is based on the \cpp{98} version.

\subsection{Linking}
\label{subs:linking}

The additional complexity compared to a simple \verb/g++/ or \verb/clang/ compilation is the linking process.

There are multiple versions of \pru{} hardware, and the generated image must map correctly to the memory sections of the \pru. Additionally the firmware loader requires a specific header file called \emph{resource table}. This resource table contains the required system resources the \pru{} uses, therefore the firmware loader can initialize these resources.

These memory regions are defined in a \verb+.cmd+ file, and necessary to compile an \elf{} image for the firmware loader.

\subsection{RPMSG library}

To utilize the advanced communication between the host processor, and the \pru, the remote processor messaging library (\emph{rpmsg}) library can be used.

This frameworks are not included in the \pru{} \cpp{} toolchain itself, but shipped with the \emph{pru-software-support-package} \citep{TI_PRUSS_REPO}. This library contains the implementation, and example usage of this library.

The \emph{rpmsg} library have two components:
\begin{itemize}
	\item A kernel module, which provides the virtual interfaces for the host application.
	\item A \cpl{} library, which the \pru{} can use to communicate with the host processor.
\end{itemize}

When a communication channel is created while loading the \pru{} code, a virtual device is created in the \emph{/dev} folder of Linux system. This virtual device can be used to communicate with the \pru{}. The usage of this \emph{/dev} is very straightforward, the virtual device is like any ordinal file. From the host application, this file can be opened, and messages can be written/read from this file. An example code is presented in \cref{lst:communication_sample}.

\begin{lstlisting}[
	float,
	caption={Example communication code using the virtual device provided by \emph{rpmsg}},
	label=lst:communication_sample,
	language={C++}
]
#include <fstream>
#include <iostream>
#include <string>

int main(int argc, const char * argv[]) {
	std::ofstream rpmsg_stream;
	rpmsg_stream.open("/dev/rpmsg_pru30");

	for(int x=0; x<10000; ++x) {
		for(int i=0; i<=6; ++i) {
			rpmsg_stream << std::to_string(i) << std::endl;
		}
	}

	return 0;
}
\end{lstlisting}

\subsubsection{Makefile}

The compilation workflow is aided by the \verb+make+ utility. While the makefile of the support library is written for Linux systems, it is compatible and was tested on \osx{} version~\verb|10.11.4|.

The \verb+make+\footnote{More information about make: \url{https://www.gnu.org/software/make/}} application is the most popular build tool in the open source community. It uses specific files called \emph{Makefile}. These files declaratively state the dependencies of each step of compilation. The advantages of using make is to:
\begin{itemize}
	\item Simplify a complex workflow, just by calling \verb+make+.
	\item The \verb+make+ utility can determine which dependency is required to compile the application. If the project e.g.\ a kernel source consists of numerous object files, one small change triggers the recompilation of the corresponding source files, and not the complete project itself.
\end{itemize}

A prepared Makefile recommended for \pru{} code compilation can be found in the repository \citep{TI_PRUSS_REPO}. The thesis uses a slightly modified version of file \url{examples/am335x/PRU_gpioToggle/Makefile} to generate the artifacts. The modifications were:
\begin{itemize}
	\item Set the mandatory \verb+PRU_CGT+ variable. This variable store the path where the toolchain binaries can be found.
	\item Define a new variable \verb+SUPPORT_PATH+. The directory structure of the original repository structure contained the support file, therefore the Makefile accessed these files with relative paths. To increase the reusability of the makefile, the \verb+SUPPORT_PATH+ stores the absolute path to the support library.
	\item Changing the \verb+LIBS+ variable to the correct directory. The original version used a relative path to the library, this change modified this by adding a local \verb+lib+ folder, and copying all the necessary \verb+.lib+ file to this directory.
	\item Changing the \verb+INCLUDE+ variable to the correct directory. To be noted, the original Makefile does not have an include directory for the project itself, because the examples are written in a single files, thus no header files were needed.
\end{itemize}

\begin{figure}[!h]
	\centering
	\begin{tikzpicture}[
		grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
		edge from parent path={[thick] (\tikzparentnode.south) |- (\tikzchildnode.west)},
		every node/.style = {
			anchor=west
		}
	]

	\node {«\,project root\,»}
		child { node {\Verb+Makefile+}}
		child { node {\Verb+AM335x\_PRU.cmd+}}
		child { node {\Verb+include+}
			child { node {«\,header files (*.hpp)\,»}}
		}
		child [missing] {}
		child { node {\Verb+src+}
			child { node {«\,source files (*.cpp)\,»}}
		}
		child [missing] {}
		child { node {\Verb+gen+}
			child { node {«\,compiled files (*.object, *.pp, *.map)\,»}}
		};

	\end{tikzpicture}
	\caption{The directory structure of a typical \pru{} application}
\label{fig:pru_dir_structure}
\end{figure}

\section{Compiling}

This section will walk through the compilation workflow, and common pitfalls.

With the provided makefile it's considerably easy to compile the application, the developer just need to execute the \verb|make all| command in the project root, where the Makefile is located. The Makefile will invoke the compiler with the required parameters.

\subsection{Compilation errors}

TODO Maybe some text here

\subsection{\cpl{} or \cpp{} error codes}

These errors are commonly associated with the syntax of the language, and not the specific compiler.

A special case is the ''\verb|#929-D: Embedded C++ does not support templates|'' error. This error is common when \verb+std+ libraries which use templating are included. This error is caused by the embedded \cpp{} flag~\cite[Section 5.13.4]{PRU_C_CPP}. This flag forbids the usage of:
\begin{itemize}
	\item Templates
	\item Exception handling
	\item Run-time type information
	\item The new cast syntax
	\item The keyword mutable
	\item Multiple inheritance
	\item Virtual inheritance
	\item \verb|iostream|
\end{itemize}

It's a good practice to use this flag. The features previously listed are the main cause of drastic code size increase, and the main bottleneck of the \pru{} is memory size.

\subsubsection{Memory overallocation issues}

The error ''\verb|#10099-D: program will not fit into available memory.| listed in \cref{lst:example_overallocation} means the source code tries to overallocate a memory regions. The two common memory region which can be overallocated is the \iram{} or \dram{}.

\begin{lstlisting}[
		float=hb,
		caption={Example of overallocation},
		label=lst:example_overallocation,
		basicstyle=\footnotesize\ttfamily
	]
"./AM335x_PRU.cmd", line 72: error #10099-D: program will not fit into
available memory.  placement with alignment fails for section ".text" size
0x21870 page 0.  Available memory ranges:
PRU_IMEM     size: 0x2000       unused: 0x1fe4       max hole: 0x1fe4
\end{lstlisting}

In the particular case of \cref{lst:example_overallocation} the \verb|.text| section\footnote{The \verb|.text| section stores the machine code of the program.} is overallocated. The required size is \verb|0x21870|\ byte even if the max hole the linker can allocate is \verb|0x1fe4|\ byte. This example was caused by the inclusion of \verb|iostream| library, with embedded \cpp{} option disabled which would prevent this.

\section{Running the firmware}

This section will focus on the firmware loader shipped with \ti{}'s Linux 4.x kernel. As were before in \cref{subsec:fw_loader_kernel4}, the 4.x loader implemented as a kernel module, and currently no userspace \api{} is provided to support the loading of a \pru{} firmware through an application. The approach this loader follows is to read out the required information from a special header linked into the \elf{} file. This header is called a \emph{resource table}, and required for starting up the specific \pru{}.

An example empty resource table (\cref{lst:empty_resource_header}) is presented to show the basis structure of a resource table. There is not much documentation for resource table description at the time this thesis was written. Example resource tables like~\cite{RESOURCE_TABLE_0}\cite{RESOURCE_TABLE_1} provides a usable base for resource headers, because it does not need to be modified on a regular basis.

\begin{lstlisting}[
	float,
	caption={Empty resource table},
	label=lst:empty_resource_header,
	language={C}
]
#ifndef _RSC_TABLE_PRU_H_
#define _RSC_TABLE_PRU_H_

#include <stddef.h>
#include <rsc_types.h>

struct my_resource_table {
	struct resource_table base;

	uint32_t offset[1]; /* Should match 'num' in actual definition */
};

#pragma DATA_SECTION(".resource_table")
#pragma RETAIN
struct my_resource_table pru_remoteproc_ResourceTable = {
	1,     /* version */
	0,     /* number of entries in the table */
	0, 0,  /* reserved, must be zero */
	0,     /* offset[0] */
};

#endif /* _RSC_TABLE_PRU_H_ */
\end{lstlisting}

After the resource table is included in the main source file and the project is compiled, the final artifact \verb|pru.out| can be found in the \verb|gen| directory. This \verb|pru.out| file is an \elf{} file, and ready to be loaded into the pru.

To make the firmwares usable by the loader, the artifact files must be copied to the \verb|/lib/firmware/| folder. Naming conventions are used to differentiate between the two \pru{}:
\begin{itemize}
	\item \verb|am335x-pru0-fw| for \pru{} core 0
	\item \verb|am335x-pru1-fw| for \pru{} core 1
\end{itemize}
Therefore if one would like to load a firmware to the \pru{} core 0, the \elf{} file must be placed under the \verb|/lib/firmware/am335x-pru0-fw| path.

TODO Console output from PRU loading
